Rapid technological progress in white light-emitting diodes \\ and its sources in innovation and technology spillovers

### 2023-08-18 2_SSL_NE/article/article_supplementary.tex
added edits

\setlength{\fboxsep}{10pt}
\fbox{
    \parbox{\textwidth}{
        \textit{\textsc{Supplementary Information}} for: \\
        \newline 
        M. P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
        \newline
        $^1$ Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, UK \\
        $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
        $^3$ Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland \\
        $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
    }
}

@@

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
            \label{fig:manuf_classical_2003}
        \end{figure}
    \end{landscape}

@@

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012. MQW - multiple quantum well; LLO - laser lift-off. Continued in \cref{fig:manuf_vtf_2012-2}}
            \label{fig:manuf_vtf_2012-1}
        \end{figure}
    \end{landscape}

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
            \caption{(2/2) Continued from \cref{fig:manuf_vtf_2012-1}. Encaps. - encapsulation.}
            \label{fig:manuf_vtf_2012-2}
        \end{figure}
    \end{landscape}

@@

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020. MQW - multiple quantum well. Continued in \cref{fig:manuf_csp_2020-2}.}
            \label{fig:manuf_csp_2020-1}
        \end{figure}
    \end{landscape}

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020. Continued from \cref{fig:manuf_csp_2020-1}. Encaps. - encapsulation.}
            \label{fig:manuf_csp_2020-2}
        \end{figure}
    \end{landscape}

@@

The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level. The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
%
\begin{equation}
\label{eqn:cost_sum}
    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
\end{equation}
%
where
%
\begin{align*}
    P_S &\dots \text{sapphire substrate price per die} \\
    C_w &\dots \text{wafer processing cost per die} \\
    C_p &\dots \text{die processing cost}
\end{align*}
%
The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
%
\begin{equation}
        C_w = \sum_i C_i
\end{equation}
%
\begin{equation}
	C_p = \sum_j C_j
\end{equation}
%
The cost of a single process step $C_i$ can now be written as
%
\begin{equation}
\label{eqn:cost_wafer}
    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
\end{equation}
%
where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
%
\begin{align*}
        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
        y &\dots \text{process step yield} \\
        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
        p &\dots \text{power consumption} \\
        c_e &\dots \text{hourly electricity cost} \\
        c_m &\dots \text{hourly maintenance cost} \\
        c_d &\dots \text{hourly depreciation cost}\\
        c_l &\dots \text{hourly labour cost} \\
        c_o &\dots \text{hourly overhead cost} \\
        t_i &\dots \text{time per run} \\
        w &\dots \text{wafers per run} \\
        v_x &\dots \text{volume of material $x$ per wafer} \\
        c_x &\dots \text{cost of material $x$ per volume}\\
\end{align*}
%
The number of die per wafer $D$ depends on the total usable wafer area. The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
%
\begin{equation}
	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
\end{equation}
%
Determining the usable wafer area as a function of these three parameters requires a numerical solution. However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
%
\begin{equation}
\label{eqn:dpw}
	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
\end{equation}
%
where
%
\begin{align*}
    d &\dots \text{wafer diameter} \\
    e &\dots \text{wafer edge exclusion zone width} \\
    a &\dots \text{die area} \\
    s &\dots \text{cutting street width} \\
\end{align*}

\cref{fig:dpw} shows the results of DPW calculations using \cref{eqn:dpw} for different wafer diameters $d$ that were most commonly used in LED manufacturing in $d(2002)=50$mm, $d(2010)=100$mm and $d(2020)=200$mm. We can see that DPW dramatically increases over time.

\begin{figure}[h!]
    \centering
    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2010)=100$mm$\rightarrow5370$DPW, $d(2020)=200$mm$ \rightarrow26,838$DPW. Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews. For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
    \label{fig:dpw}
\end{figure}

@@

\subsubsection{Example of Input Data Considered: Sapphire Wafers}

@@

Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing (see top panel of \cref{fig:wafers}). The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}. We use the market average for the sapphire wafer price in different years as a direct input into the cost model, and the dominant wafer diameter (see bottom panel in \cref{fig:wafers}) in different years as an input to calculate DPW in \cref{eqn:dpw}.

\begin{figure}[h!]
    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
	\caption{Historical data for sapphire substrate prices of different surface properties and diameters. Shown is data for polished surface (NPS) and patterned substrates (PSS). \textit{Monocrystal} denotes the Russian manufacturer of the same name. Dashed lines are projections from the previous year. Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}. Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes. Hatched bars are projections. Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
	\label{fig:wafers}
\end{figure}

@@

To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}. Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
%
\begin{equation}
F=ab+cd
\end{equation}
%
and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
%
\begin{equation}
\Delta F = F(t_1)-F(t_0) = \sum_{i=a, \dots, d} \Delta F_i
\end{equation}
%
Index decomposition analysis has been developed to solve this problem \cite{Boyd1987}. A short derivation, based on \cite{kavlak2018evaluating} is provided here:

The infinitesimal contribution to the total function value by the infinitesimal change in an input variable is defined through the total differential of the function as
%
\begin{equation}
\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
\end{equation}
%
where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
%
\begin{equation}
\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
\label{eqn:integral_1}
\end{equation}
%
However, data on the input variables in our study is often not available in continuous time. Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model. This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}. In our work, we propose to address this problem by following an approach developed by Kavlak et al. \cite{kavlak2018evaluating}. For the detailed derivation, we refer to this publication.

The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
%
\begin{equation}
    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
\end{equation}
where
\begin{equation}
    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
\end{equation}
%
which can be written as a constant part $F$ and a part describing the dependence of the $i$-th cost component on the $w$-th variable.
%
Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
\begin{equation}
    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
\end{equation}
%
where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$. In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$ \cite{kavlak2018evaluating}. The contribution of a single cost model variable $r_z$ can then be written as
%
\begin{equation}
    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
\end{equation}
%
We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article. Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.

@@

Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}. Specifically, the figure includes data for forward voltage used to calculate to forward voltage efficiency $\eta_{V_f}$ (Panel A1); efficiency droop $\eta_droop$ (Panel A2); internal quantum efficiency $\eta_IQ$ (Panel B1); and light extraction efficiency $\eta_LE$ (Panel B2). Additional information and context for individual panels in Supplementary \cref{fgr:subeff} are provided below.

@@

\textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used. Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}. Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}. “Not reported” denotes data points where the IQE measurement technique was not reported. Whiskers indicate reported ranges, where applicable. \

\textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency. The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given. Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr. - electrode; text. - textured.

Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$. Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the empirical \cref{fgr:subeff} not relevant. Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum}.

\begin{figure}[h!]
\centering
\includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
\caption{Historical improvements in light-emitting diode technology. Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$. Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$. Source: own synthesis of data from the full list of sources provided in \cref{sec:sources}. Data on average device performance adapted from U.S. Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
\label{fgr:subeff}
\end{figure}

@@

\begin{table}[h!]
    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies. The table includes innovations and technology improvements affecting forward voltage efficiency, light extraction efficiency, internal quantum efficiency, and light conversion efficiency. For a list of innovations affecting spectral efficiency, see Table 1 in the main article.}

@@

        \multicolumn{5}{c}{Forward Voltage Efficiency (VFE) $\eta_{Vf}$} \\

@@

            1994 & Double heterostructure & Higher Radiative recombination probability & No & \cite{Nakamura1994} \\

@@

            1996 & Multiple quantum well & Higher Radiative recombination probability & No & \cite{Koike1996} \\

@@

            ongoing & Lower Current Density & Current Density & No & I \\

@@

            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{bhardwaj2016progress} \\

@@

            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{Wildeson2017} \\

@@

            $<$ 2017 & Defect Getting \newline Underlayer & Defect Density & No & I, \cite{haller2017burying} \\

@@

        \multicolumn{5}{c}{Light Conversion efficiency (CE) $\eta_{C}$} \\

@@

            $<$ 2017 & Defect gettering \newline underlayer & Defect Density & No & I, \cite{haller2017burying} \\

@@

Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 6 in the main article). An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG). Used in combination with red-emitting sulfide phosphors, by 2002 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}. However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased. New chemically stable and non-toxic red phosphors were needed.

>>>>

\setlength{\fboxsep}{10pt}
\fbox{
    \parbox{\textwidth}{
        \textit{\textsc{Supplementary Information}} for: \\
        \newline 
        M.P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
        \newline
        $^1$ Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, UK \\
        $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
        $^3$ Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland \\
        $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
    }
}

@@

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
            \caption{\textbf{Manufacturing process for a classical LED package with lateral current spreading, circa 2003.} MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
        \label{fig:manuf_classical_2003}
        \end{figure}
    \end{landscape}

@@

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
            \caption{\textbf{Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012. (1/2)} MQW - multiple quantum well; LLO - laser lift-off. Continued in \cref{fig:manuf_vtf_2012-2}}
            \label{fig:manuf_vtf_2012-1}
        \end{figure}
    \end{landscape}

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
            \caption{\textbf{Manufacturing process for a vertical thin-film package flip-chip LED chip with vertical current spreading, circa 2012 (2/2)}. Continued from \cref{fig:manuf_vtf_2012-1}. Encaps. - encapsulation.}
            \label{fig:manuf_vtf_2012-2}
        \end{figure}
    \end{landscape}

@@

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020. (1/2)} MQW - multiple quantum well. Continued in \cref{fig:manuf_csp_2020-2}.}
            \label{fig:manuf_csp_2020-1}
        \end{figure}
    \end{landscape}

    \begin{landscape}
        \begin{figure}
            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020. (2/2)}. Continued from \cref{fig:manuf_csp_2020-1}. Encaps. - encapsulation.}
            \label{fig:manuf_csp_2020-2}
        \end{figure}
    \end{landscape}

@@

The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level. The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
%
\begin{equation}
\label{eqn:cost_sum}
    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
\end{equation}
%
where
%
\begin{align*}
    P_S &\dots \text{sapphire substrate price per die} \\
    C_w &\dots \text{wafer processing cost per die} \\
    C_p &\dots \text{die processing cost}
\end{align*}
%
The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
%
\begin{equation}
        C_w = \sum_i C_i
\end{equation}
%
\begin{equation}
	C_p = \sum_j C_j
\end{equation}
%
The cost of a single process step $C_i$ can now be written as
%
\begin{equation}
\label{eqn:cost_wafer}
    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
\end{equation}
%
where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
%
\begin{align*}
        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
        y &\dots \text{process step yield} \\
        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
        p &\dots \text{power consumption} \\
        c_e &\dots \text{hourly electricity cost} \\
        c_m &\dots \text{hourly maintenance cost} \\
        c_d &\dots \text{hourly depreciation cost}\\
        c_l &\dots \text{hourly labour cost} \\
        c_o &\dots \text{hourly overhead cost} \\
        t_i &\dots \text{time per run} \\
        w &\dots \text{wafers per run} \\
        v_x &\dots \text{volume of material $x$ per wafer} \\
        c_x &\dots \text{cost of material $x$ per volume}\\
\end{align*}
%
The number of die per wafer $D$ depends on the total usable wafer area $A_{\text{usable}}$. The usable area depends on the total area of the wafer $A_{\text{wafer}}$ , the cutting street width area between the chips $A_{\text{cut}}$  and the exclusion zone at the rim of the wafer $A_{\text{exclusion}}$ .
%
\begin{equation}
	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
\end{equation}
%
Determining the usable wafer area as a function of these three parameters requires a numerical solution. However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
%
\begin{equation}
\label{eqn:dpw}
	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
\end{equation}
%
where
%
\begin{align*}
    d &\dots \text{wafer diameter} \\
    e &\dots \text{wafer edge exclusion zone width} \\
    a &\dots \text{die area} \\
    s &\dots \text{cutting street width} \\
\end{align*}

\cref{fig:dpw} shows the results of DPW calculations using \cref{eqn:dpw} for different wafer diameters $d$ that were most commonly used in LED manufacturing in 2003 (50mm), 2010 (100mm) and 2020 (200mm). We can see that DPW dramatically increases over time.

\begin{figure}[h!]
    \centering
    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
    \caption{\textbf{Number of die per wafer for the dominant wafer diameters used in different years.} $d(2003)=50$mm$\rightarrow851$ DPW, $d(2010)=100$mm$\rightarrow5370$ DPW, $d(2020)=200$mm$\rightarrow26,838$ DPW. Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews. For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
    \label{fig:dpw}
\end{figure}

@@

\subsubsection{Input Data on Sapphire Wafers}

@@

Wafer sizes used in manufacturing have also increased over the years due to favourable economics of large wafer processing (see bottom panel in \cref{fig:wafers}). The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}. We use the market average for the sapphire wafer price in different years as a direct input into the cost model, and the dominant wafer diameter (see bottom panel in \cref{fig:wafers}) in different years as an input to calculate DPW in \cref{eqn:dpw}.

\begin{figure}[h!]
    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
	\caption{\textbf{Historical trends in sapphire wafer size and price.} Top panel: Historical data for sapphire substrate prices of different surface properties and diameters, 2004-2020. Shown is data for non-patterned (polished) surface (NPS) and patterned surface substrates (PSS). Monocrystal denotes the Russian manufacturer of the same name. Dashed lines are projections made in 2015. Sources: (Monocrystal 2020)(Yole Developpement 2011)(Yole Developpement 2015). Bottom panel: Prevalence of sapphire wafer size used in LED manufacturing in 2008-2020. Hatched bars are projections. Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
	\label{fig:wafers}
\end{figure}

@@

To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}. Mathematically, given a function $F$ which describes the manufacturing cost or performance function of a device at time $t$,
%
\begin{equation}
F=ab+cd
\end{equation}
%
and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
%
\begin{equation}
\Delta F = F(t_1)-F(t_0) = \sum_{i=a, \dots, d} \Delta F_i
\end{equation}
%
Index decomposition analysis has been developed to solve this problem \cite{Boyd1987}. A short derivation, based on \cite{kavlak2018evaluating} is provided below.

The infinitesimal contribution to the total function value by the infinitesimal change in an input variable is defined through the total differential of the function as
%
\begin{equation}
\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
\end{equation}
%
where $x_i$ is an input variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
%
\begin{equation}
\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
\label{eqn:integral_1}
\end{equation}
%
However, data on the input variables in our study is often not available in continuous time. Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model. This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}. In our work, we propose to address this problem by following an approach developed by Kavlak et al. \cite{kavlak2018evaluating}. For the detailed derivation, we refer to this publication.

In this approach, the function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
%
\begin{equation}
    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
\end{equation}
where
\begin{equation}
    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
\end{equation}
%
which can be written as a constant part $F$ and a part describing the dependence of the $i$-th cost component on the $w$-th variable.
%
Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
\begin{equation}
    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
\end{equation}
%
where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$. In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$ \cite{kavlak2018evaluating}. The contribution of a single cost model variable $r_z$ can then be written as
%
\begin{equation}
    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
\end{equation}
%
We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in relation to Figure 5 in the main article. Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.

@@

Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}. Specifically, the figure includes data for forward voltage used to calculate to forward voltage efficiency $\eta_{V_f}$ (Panel A1); efficiency droop $\eta_{droop}$ (Panel A2); internal quantum efficiency $\eta_{IQ}$ (Panel B1); and light extraction efficiency $\eta_{LE}$ (Panel B2). Additional information and context for individual panels in \cref{fgr:subeff} are provided below.

@@

\textbf{Panel B1:} Internal quantum efficiency (IQE) for different chip architectures by type of measurement used. Note that the bump in the DOE Average around 2013 is due to a change in IQE definition laid out in \cite{doe_ssl_multiyear_2013}. Measurement methods: PL - Photoluminescence \cite{Shim_2018}, EL - Electroluminescence \cite{Getty_2009}; ABC Model \cite{Karpov_2014}, Ideality Factor \cite{Masui_2010}. “Not reported” denotes data points where the IQE measurement technique was not reported. Whiskers indicate reported ranges, where available. \

\textbf{Panel B2:} Light extraction efficiency (LEE) for different chip architectures by type of specific technology used to improve LEE. The types of data points are shown for reference: Simulation - ray-tracing computer simulation; Experimental (absolute) – direct LEE values; Experimental (relative) - relative improvement over baseline chip architecture. Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr. - electrode; text. - textured.

Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.  Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant. Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).

\begin{figure}[h!]
\centering
\includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
\caption{\textbf{Historical improvements in LED sub-efficiencies.} Panels A1-A2 show data for key metrics used to calculate forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$. Panels B1-B2 show data for internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$. Source: own synthesis of data from the list of sources provided in \cref{sec:sources}. Data on average device performance adapted from U.S. Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
\label{fgr:subeff}
\end{figure}

@@

\begin{table}[H]
    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.}

@@

        \multicolumn{5}{c}{Forward Voltage Efficiency (FVE) $\eta_{Vf}$} \\

@@

            1994 & Double heterostructure & Higher radiative recombination probability & No & \cite{Nakamura1994} \\

@@

            1996 & Multiple quantum well & Higher radiative recombination probability & No & \cite{Koike1996} \\

@@

            ongoing & Lower current Density & Current density & No & I \\

@@

            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\

@@

            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{Wildeson2017} \\

@@

            $<$ 2017 & Defect getting \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\

@@

        \multicolumn{5}{c}{Light conversion efficiency (CE) $\eta_{C}$} \\

@@

            $<$ 2017 & Defect gettering \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\

@@

Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 6 in the main article). An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG). Used in combination with red-emitting sulfide phosphors, by 2003 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2003}. However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased. New chemically stable and non-toxic red phosphors were needed.
